---
title: "Exploratory Data Analysis in R for Data Science"
output:
  html_document:
    df_print: paged
---

#EDA = "exploratory data analysis"
# can't knit because "there is no package called 'viridisLite'"

```{r}
library(tidyverse)
```

# our goal is to ask quaility questions (not just a lot of them)
# 2 ex. of Q are:
#1) What type of variation occurs within my variables?
#2) What type of covariation occurs between my variables?

## Use bar chart w/ categorical data
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut)) #creates boxplot of diaminds based on cut
diamonds %>% 
  count(cut) #number chart of diamond cut
diamonds %>% 
  count(cut_width(carat, 0.5)) #number of diamonds w ea cut width
smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1) #adjust binwidth to reveal diff patterns

ggplot(data = smaller, mapping = aes(x = carat, colour = cut))+
  geom_freqpoly(binwidth = 0.1) #displays cuts as overlapping lines on chart
# ^ "Error in loadNamespace(name) : there is no package called ‘viridisLite’". I tried installing and calling up this 
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual #pluck out unusual data outside of limits
```
##7.3.4
#Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth. **Length & width are probably similar and depth is the x value since it is the smallest and diamonds aren't that deep**
```{r}
ggplot(data = diamonds) + 
       geom_histogram(mapping = aes(x = x), binwidth = 0.01) #most values fall bw 4-9
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = x)) #hundreds of counts, most ~5
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = y)) #all values until 10, some w 100s of counts
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = z)) #looks like y data
```


#Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.) The better quality diamonds are cheaper on average while the "fair" omes cost more

#How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?
```{r}
diamonds %>% 
  count(carat=.99, carat=1)
```


#Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = y), xlim(10))
```


```{r}
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y)) #replaces the unusual values with missing values "NA"
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE) #na.rm = TRUE supresses "missing values" banner

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4) 
```
#What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference? They are not included in a bar chart while a hisotgram will not appear unless the values are accounted for. 
```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_histogram(na.rm = TRUE)
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_bar(na.rm = TRUE)
```

#What does na.rm = TRUE do in mean() and sum()?
```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(mean(na.rm = TRUE)) # argument "x" is missing, with no default

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()+
 sum(na.rm = TRUE) #"Don't know how to add sum(na.rm = TRUE) to a plot"
```
#Covariation = the tendency for the values of two or more variables to vary together in a related way
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))#reorders class based on the median value of hwy
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip() #good for long varible names
```
#Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.
```{r}
?nycflights13::flights
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(x = sched_dep_time, y = carrier)) + 
    geom_point(mapping = aes(colour = cancelled))
```

#What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?
```{r}

```
#Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?
```{r}
install.packages(ggstance) #object 'ggstance' not found
```
#One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs cut. What do you learn? How do you interpret the plots?
```{r}
install.packages(lvplot) #object 'lvplot' not found
```

#Compare and contrast geom_violin() with a facetted geom_histogram(), or a coloured geom_freqpoly(). What are the pros and cons of each method?
```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
geom_violin(na.rm = TRUE) #trippy, cool graph that looks like an extra warped graph
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
geom_histogram(na.rm = TRUE) #stat_bin() must not be used with a y aesthetic
```


#If you have a small dataset, it’s sometimes useful to use geom_jitter() to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to geom_jitter(). List them and briefly describe what each one does.
```{r}
??ggbeeswarm #can create violin graph w/ individual data dots; dodging can be used to compare within groups
#pseudorandom =  kernel density is estimated then points are distributed uniform randomly within the density estimate for a given bin
#alternating within bins = the kernel density is estimated then points are distributed within the density estimate for a given bin evenly spaced with extreme values alternating from R to L
#tucky = strips displaying empirical distributions
#quasirandom The kernel density is estimated then points are distributed quasirandomly using the von der Corput sequence within the density estimate for a given bin
```
```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n)) #blue pixels graph
```
#How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut? Increase the number of boxes.

#Use geom_tile() together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it? Thus data is very crowded.
```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(x = sched_dep_time, y = carrier)) + 
    geom_tile(mapping = aes(colour = cancelled))
```


#Why is it slightly better to use aes(x = color, y = cut) rather than aes(x = cut, y = color) in the example above? The cateogorical varible works better as color while the quantitative one is the y.
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, , alpha = 1 / 100)
#2 categorical variables w/ exponential relationship
#transparency helps w big data sets because points begin to overplot

ggplot(data = smaller) +
  geom_bin2d(mapping = aes(x = carat, y = price))

install.packages(hexbin)
ggplot(data = smaller) +
  geom_hex(mapping = aes(x = carat, y = price)) #hexagon exponential trend

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1))) #many boxplots
```
#Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?

#How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you? Small diamonds cost more on average.

#Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.
```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1), price)) 
```


#Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately.
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11)) #clear trend shows
```
##Patterns & models
#If you spot a pattern, ask yourself:
#Could this pattern be due to coincidence (i.e. random chance)?
#How can you describe the relationship implied by the pattern?
#How strong is the relationship implied by the pattern?
#What other variables might affect the relationship?
#Does the relationship change if you look at individual subgroups of the data?
```{r}
ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))# old faithful scatterplot

library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid)) #remove the strong relationship between carat and price to see that, relative to their size, better quality diamonds are more expensive

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))

ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))
```
#Long and short code
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_freqpoly(binwidth = 0.25) # explicit version

ggplot(faithful, aes(eruptions)) + 
  geom_freqpoly(binwidth = 0.25) # consise version

diamonds %>% 
  count(cut, clarity) %>% 
  ggplot(aes(clarity, cut, fill = n)) + 
    geom_tile() #note the transition from the pip to the +
```

